\proglang{Stan}: A Probabilistic Programming Language

### 2014-12-01 src/docs/papers/stan-language/article.tex
updated article for reviewer comments

@@

  \proglang{Stan} can be called from the command line, through
  \proglang{R} using the \pkg{RStan} package, or through
<  \proglang{Python} using the \pkg{PyStan} package.  All three
  interfaces support sampling or optimization-based inference and
  analysis, and \pkg{RStan} and \pkg{PyStan} also provide access to
  log probabilities, gradients, Hessians, and I/O transforms.
}

@@

\section[Why Stan?]{Why \proglang{Stan}?}
%% Note: If there is markup in \(sub)section, then it has to be escape as above.

We did not set out to build \proglang{Stan} as it currently exists.%
%
\footnote{\proglang{Stan}'s home page, \url{http://mc-stan.org/},
  links to the \proglang{Stan} manual, example models including translations of
  the \proglang{BUGS} examples and the models from \citep{GelmanHill:2007},
  getting started instructions and full documentation for \proglang{Stan}'s
  interfaces for the command line shell (\pkg{CmdStan}), \proglang{Python}
  (\pkg{PyStan}), and \proglang{R} (\pkg{RStan}), and the source code repositories
  and issue trackers.}
%
Our original goal was to apply full Bayesian inference to the sort of multilevel
generalized linear models discussed in Part~II of
\citep{GelmanHill:2007}, which are structured with grouped and
interacted predictors at multiple levels, hierarchical covariance
priors, nonconjugate coefficient priors, latent effects as in
item-response models, and varying output link functions and
distributions.

These models turned out to be a challenge for existing general purpose
inference software.  A direct encoding in \proglang{BUGS}
\citep{LunnEtAl:2000,LunnEtAl:2009,LunnEtAl:2012} or
\proglang{JAGS}~\citep{Plummer:2003} can grind these tools to a halt.
We began by adding custom vectorized logistic regressions to
\proglang{JAGS} using \proglang{C++} to overcome the cost of
interpretation.  Although this is much faster than looping in
\proglang{JAGS}, it quickly became clear that the root of the problem
was the slow convergence of conditional sampling when parameters were
highly correlated in the posterior, as for time-series models and
hierarchical models with interacted predictors.  We finally realized
we needed a better sampler, not a more efficient implementation of
Gibbs sampling.

We briefly considered trying to tune proposals for a random-walk
Metropolis-Hastings sampler, but that seemed too problem specific and
not even necessarily possible without some kind of adaptation rather
than a global tuning of the proposals.

We were hearing more about Hamiltonian Monte Carlo (HMC) sampling,
which appeared promising but was also problematic in that the
Hamiltonian dynamics simulation requires the gradient of the log
posterior.  Although not difficult in theory for most functions,
computing these gradients by hand on a model-by-model basis is very
tedious and error prone.  That is when we discovered reverse-mode
algorithmic differentiation, which, given a templated \proglang{C++}
function for the log posterior, automatically computes its analytic
gradient up to machine precision accuracy in only a few multiples of
the time to evaluate the log probability function itself.%
%
\footnote{Hessian matrices (i.e., all second order derivatives) are
  more expensive to calculate; each row corresponds to a parameter and
  is filled by the gradient of the derivative of the log probability
  function with respect to the parameter.}
%
We explored existing algorithmic differentiation packages with open
licenses such as \pkg{RAD} \citep{Gay:2005} and its repackaging in the
Sacado module of the \pkg{Trilinos} toolkit, and the \pkg{CppAD}
package of the COIN-OR toolkit \citep{BellBurke:2008}.  Both packages,
however, supported few special functions (e.g., probability functions,
log gamma, inverse logit) or linear algebra operations (e.g., Cholesky
decompositions, matrix division), and neither are easily or modularly
extensible.

Consequently we built our own reverse-mode algorithmic differentiation package.
At that point, we ran into the problem that we could not just plug in
the probability functions from a package like \pkg{Boost} because they
weren't templated generally enough across all arguments.  Rather
than pay the price of promoting floating point values to algorithmic
differentiation variables, we wrote our own fully templated
probability functions and other special functions.

Next, we integrated the \proglang{C++} package \pkg{Eigen} for matrix
operations and linear algebra functions.  \pkg{Eigen} makes extensive
use of expression templates for lazy evaluation and the curiously
recurring template pattern (CRTP) to implement concepts without
virtual function calls; \cite{VandevoordeJosuttis:2002} provide a
complete description of template metaprogramming techniques.
Unfortunately, we ran into the same problem with Eigen as with the
existing probability libraries---it doesn't support mixed operations
of algorithmic differentiation variables and primitives like
\code{double}.  Although we initially began by promoting
floating-point vectors to algorithmic differentiation variables, we
later completely rewrote mixed-mode operations like multiplying a data
matrix by a parameter vector.  To compute derivatives of matrix
operations such as division and inverse, we implemented specialized
derivatives using the rules described by \cite{Giles:2008}.

At this point, we could fit models coded directly in \proglang{C++} on
top of the pre-release versions of the \proglang{Stan} application
programming interface (API).  Seeing how well this all worked, we set
our sights on the generality and ease of use of \proglang{BUGS} and
designed a modeling language in which statisticians could write their
models in familiar notation that could be transformed to efficient
\proglang{C++} code before being compiled into an efficient executable
program.  Although we started with the target of specifying directed
graphical models, our first implementation turned out to be a more
general imperative form of specifying log probability functions;
\proglang{BUGS} models could be translated line for line, but the
\proglang{Stan} language also supports much richer imperative
constructs like conditionals, while loops, and local variables.  This
paper is primarily about the \proglang{Stan} language and how it can
be used to specify statistical models.

The next problem we ran into when we started implementing richer
models was variables with constrained support, such as positive variables, 
simplexes, and covariance matrices.  
Efficient implementation of these constraints
required the introduction of typed variables which automatically transformed to
unconstrained support with suitable adjustments to the log probability
from the log absolute Jacobian determinant of the inverse transforms.

Even with the prototype compiler generating models, we still faced a
major hurdle for ease of use. The efficiency of HMC is very sensitive
to two tuning parameters, the discretization interval (i.e., step
size) and the total simulation time (i.e., number of steps).  The
interval size parameter could be tuned during warm-up based on
Metropolis rejection rates, but the number of steps proved difficult
to tune without sacrificing the detailed balance of the sampler.  This
led to the development of the No U-Turn (NUTS) sampler
\citep{HoffmanGelman:2011}.  Roughly speaking, NUTS builds a tree of
possible samples by randomly simulating Hamiltonian dynamics both
forwards and backwards in time until the combined trajectory turns
back on itself.  Once the trajectory has terminated, a new sample is
drawn from the tree.  In its original version, NUTS also estimates a
step size during warmup based on a target acceptance probability.

We thought we were home free, but when we measured the
speed of some \proglang{BUGS} examples versus \proglang{Stan} we were
very disappointed.  \proglang{BUGS}'s very first example model, Rats,
ran more than an order of magnitude faster in \proglang{JAGS} than in
\proglang{Stan}.  Rats is a tough test case because the conjugate
priors and lack of posterior correlations make it an ideal candidate
for efficient Gibbs sampling.  Realizing that we were doing redundant
calculations, we wrote a vectorized form of the normal distribution
for multiple variates with the same mean and scale, which sped things
up a bit.  Performance was lacking until we finally figured out how
to both eliminate redundant calculations and partially evaluate the
gradients using a combination of expression templates and
metaprogramming.

When we attempted to fit a time-series model, we found that
normalizing the data to unit sample mean and variance sped up the fits
by an order of magnitude.  In hindsight this isn't surprising, as the
performance of the numerical simulation in HMC scales with the
variation of the parameter scales.  In \proglang{Stan} 1.0 we
introduced an adaptive diagonal metric (mass matrix) into our HMC
implementations that allowed the parameter scales to be standardized
automatically; \proglang{Stan} 2.0 added an option to estimate a dense metric
(mass matrix) and a full covariance standardization of the posterior.

Because the diagonal and dense metrics perform only a global
standardization, the performance of \proglang{Stan} in models where
the relative parameter scales varies by location, such as hierarchical
and latent models, can suffer (hierarchical modeling is covered in
Section~\ref{hierarchical-modeling.section} and reparameterizations to speed it
up in Section~\ref{centering.section}).  Riemannian manifold Hamiltonian Monte
Carlo (RHMC) introduces a location-dependent metric that can overcome
these final hurdles \citep{GirolamiCalderhead:2011}; \proglang{Stan} has a
prototype implementation of RHMC based on the SoftAbs metric of
\citep{Betancourt:2012}, using a generalization of NUTS to Riemannian
manifolds \citep{Betancourt:2013}.

\section{Overview}

@@

Consider estimating the chance of success parameter for a Bernoulli
distribution based on a sequence of observed binary outcomes.  
Figure~\ref{bernoulli-model.fig} provides an implementation of such a
model in \proglang{Stan}.%
%
\footnote{This model is available in the \proglang{Stan} source
  distribution in \nolinkurl{src/models/basic\_estimators/bernoulli.stan}.}
%
\begin{figure}
\begin{Code}
data {
  int<lower=0> N;                  // N >= 0
  int<lower=0,upper=1> y[N];       // y[n] in { 0, 1 }
}
parameters {
  real<lower=0,upper=1> theta;     // theta in [0, 1]
}
model {
  theta ~ beta(1,1);               // prior
  y ~ bernoulli(theta);            // likelihood
}
\end{Code}
\caption{\it Model for estimating a Bernoulli parameter.}\label{bernoulli-model.fig}
\end{figure}
%
The model treats the observed binary data, \code{y[1],...,y[N]}, as
independent and identically distributed, with success probability
\code{theta}.  The vectorized likelihood statement can also be coded
using a loop as in \proglang{BUGS}, although it will run more slowly than the
vectorized form:
%
\begin{Code}
for (n in 1:N)
  y[n] ~ bernoulli(theta);
\end{Code}
%
A \code{beta(1,1)} (i.e., uniform) prior is placed on \code{theta},
although there is no special behavior for conjugate priors in
\proglang{Stan}.  The prior could be dropped from the model altogether
because parameters start with uniform distributions on their support,
here constrained to be between 0 and 1 in the parameter declaration
for \code{theta}.

@@

Data for running \proglang{Stan} from the command line can be included
in \proglang{R} dump format.%
%
\footnote{A \pkg{JSON} interface for structured data input and output is
  currently under development.}
%
For example, 10 observations for the
model in Figure~\ref{bernoulli-model.fig} could be encoded as%
%
\footnote{This data
file is provided with the \proglang{Stan} distrbution in file
\nolinkurl{src/models/basic\_estimators/bernoulli.R.stan}.}
%
\begin{Code}
N <- 10
y <- c(0,1,0,0,0,0,0,0,0,1)
\end{Code}
%
This defines the contents of two variables, an integer \code{N} and a
10-element integer array \code{y}.  The variable \code{N} is declared
in the data block of the program as being an integer greater than or
equal to zero;  the variable \code{y} is declared as an integer array of size
\code{N} with entries between 0 and 1 inclusive.

@@

The executable can be run with default options by specifying a path to
the data file.  The first command in the following example changes the
current directory to that containing the model, which is where the
data resides and where the executable is built.  From there, the path
to the data is just the file name \code{bernoulli.data.R}.
%
\begin{CodeChunk}
\begin{CodeInput}
> cd $stan/src/models/basic_estimators
> ./bernoulli sample data file=bernoulli.data.R
\end{CodeInput}
\end{CodeChunk}
%
For Windows, the \code{./} before the command should be removed.

@@

  Elapsed Time: 0.00932 seconds (Warm-up)
                0.016889 seconds (Sampling)
                0.026209 seconds (Total)
\end{Code}
%
The sampler configuration parameters are echoed, here they are all
default values other than the data file.  These parameters may be set
on the command line.

@@

The output CSV file, written by default to \code{output.csv}, starts
with a summary of the configuration parameters for the run.
%
\begin{Code}
# stan_version_major = 2
# stan_version_minor = 1
# stan_version_patch = 0
# model = bernoulli_model
# method = sample (Default)
#   sample
#     num_samples = 1000 (Default)
#     num_warmup = 1000 (Default)
#     save_warmup = 0 (Default)
#     thin = 1 (Default)
#     adapt
#       engaged = 1 (Default)
#       gamma = 0.050000000000000003 (Default)
#       delta = 0.80000000000000004 (Default)
#       kappa = 0.75 (Default)
#       t0 = 10 (Default)
#       init_buffer = 75 (Default)
#       term_buffer = 50 (Default)
#       window = 25 (Default)
#     algorithm = hmc (Default)
#       hmc
#         engine = nuts (Default)
#           nuts
#             max_depth = 10 (Default)
#         metric = diag_e (Default)
#         stepsize = 1 (Default)
#         stepsize_jitter = 0 (Default)
# id = 0 (Default)
# data
#   file = bernoulli.data.R
# init = 2 (Default)
# random
#   seed = 847896134
# output
#   file = output.csv (Default)
#   diagnostic_file =  (Default)
#   refresh = 100 (Default)
\end{Code}
%
\proglang{Stan}'s behavior is fully specified by these configuration
parameters.  By using the same version of \proglang{Stan} and these
configuration parameters, exactly the same output file can be
reproduced.  The pseudorandom numbers generated by the sampler are
fully determined by the seed (here randomly generated based on the
time of the run, with value \code{847896134}) and the identifier (here
\code{0}).  The identifier is used to advance the underlying
pseudorandom number generator a sufficient number of values that
using multiple chains with the same seed and different identifiers
will draw from different subsequences of the pseudorandom number
stream determined by the seed.

The output contiues with a CSV header naming the columns of the
output. For the default NUTS sampler in \proglang{Stan} 2.2.0, these are
%
\begin{Code}
lp__,accept_stat__,stepsize__,treedepth__,n_divergent__,theta
\end{Code}
%
The values for \code{lp\_\_} indicate the log probability (up to an
additive constant).  The column headed by \code{accept\_stat\_\_}
provides the Metropolis/slice acceptance statistic for each
iteration.%
%
\footnote{Acceptance is the usual notion for a Metropolis sampler such
  as HMC \citep{MetropolisEtAl:1953}.  For NUTS, the acceptance
  statistic is defined as the average acceptance probabilities of all
  possible samples in the proposed tree; NUTS itself uses a slice
  sampling algorithm for rejection \citep{Neal:2003, HoffmanGelman:2011}.}
%
The column \code{stepsize\_\_} indicates the step size
(i.e., time interval) of the simulated trajectory, while the column 
\code{treedepth\_\_} gives the tree depth for NUTS, defined as the 
log base 2 of the total number of steps in the trajectory.
The rest of the header will be the names of parameters; in this
example, \code{theta} is the only parameter.

@@

\section{Models}

\subsection{Example: hierarchical model, with inference}\label{hierarchical-modeling.section}

\cite[Section 5.1]{GelmanEtAl:2013} define a hierarchical model of the
incidence of tumors in rats in control groups across trials; a very
similiar model is defined for mortality rates in pediatric surgeries
across hospitals in \citep[Examples, Volume
1]{LunnEtAl:2000,LunnEtAl:2009}.  
%  
\begin{figure}
\begin{Code}
data {
  int<lower=0> J;                         // number of items
  int<lower=0> y[J];                      // number of successes for j
  int<lower=0> n[J];                      // number of trials for j
}
parameters {
  real<lower=0,upper=1> theta[J];        // chance of success for j
  real<lower=0,upper=1> lambda;          // prior mean chance of success
  real<lower=0.1> kappa;                 // prior count
}
transformed parameters {
  real<lower=0> alpha;                   // prior success count
  real<lower=0> beta;                    // prior failure count
  alpha <- lambda * kappa;
  beta <- (1 - lambda) * kappa;
}
model {
  lambda ~ uniform(0,1);                 // hyperprior
  kappa ~ pareto(0.1,1.5);               // hyperprior 
  theta ~ beta(alpha,beta);              // prior
  y ~ binomial(n,theta);                 // likelihood
}
generated quantities {
  real<lower=0,upper=1> avg;             // avg success
  int<lower=0,upper=1> above_avg[J];     // true if j is above avg
  int<lower=1,upper=J> rnk[J];           // rank of j
  int<lower=0,upper=1> highest[J];          // true if j is highest rank
  avg <- mean(theta);
  for (j in 1:J)
    above_avg[j] <- (theta[j] > avg);
  for (j in 1:J) {
    rnk[j] <- rank(theta,j) + 1;
    highest[j] <- rnk[j] == 1;
  }
}
\end{Code}
  \caption{\it Hierarchical binomial model with posterior inferences,
    coded in \proglang{Stan}.}\label{hier-binom.fig}
\end{figure}
%
A \proglang{Stan} implementation is provided in
Figure~\ref{hier-binom.fig}.  In the rest of this section, we will
walk through what the meaning of the various blocks are for the
execution of the model.

The transformed data block starts with a sequence of variable
declarations and continues with a sequence of statements defining the
variables.  For example, the following transformed data block declares
a vector \code{x\_std}, then defines it to be the standardization of \code{x}.
%
\begin{Code}
transformed data {
  vector[N] x_std;
  x_std <- (x - mean(x)) / sd(x);
}
\end{Code}

The parameter block in the program in Figure~\ref{hier-binom.fig}
defines three parameters.  The parameter \code{theta[j]} represents
the probability of success in group \code{j}.  The prior on each
\code{theta[j]} is parameterized by a prior mean chance of success
\code{lambda} and prior count \code{kappa}.  Both \code{theta[j]} and
\code{lambda} are constrained to fall between zero and one, whereas
\code{lambda} is constrained to be greater than or equal to 0.1 to
match the support of the Pareto hyperprior it receives in the model block.

The parameter block is executed every time the log probability is
evaluated.  This may be multiple times per iteration of a sampling or
optimization algorithm.  Furthermore, different samplers and
optimizers use different instantiations of the log probability
function depending on the form and order of the derivative information
they require; see Section~\ref{inference-engines.section} for
details.

The transformed parameter block is executed after the parameter
block.  Constraints are validated after all of the statements defining the
transformed parameters have executed.  Failure to validate a
constraint results in an exception being thrown, which halts the
execution of the log probability function.  The log probability
function can be defined to return negative infinity or a special
not-a-number value.

If transformed parameters are used on the left-hand side of a sampling
statement, it is up to the user to add the appropriate log Jacobian
adjustment to the log probability accumulator.  For instance, a
lognormal variate could be generated as follows without the built-in
\code{lognormal} density function  using the normal density as
%
\begin{Code}
parameters {
  real<lower=0> u;
  ...
transformed parameters {
  real v;
  v <- log(u);
  increment_log_prob(u);      // log absolute Jacobian adjustment
}
model {
  v ~ normal(0,1);
}
\end{Code}
%
The transorm is $f(u) = \log u$, the inverse transform is $f^{-1}(v) =
\exp v$, so the absolute log Jacobian is $|\frac{d}{dv} \exp v| =
\exp v = u$.  Whenever a transformed parameter is used on the left
side of a sampling statement, a warning is printed to remind the user
of the need for a Jacobian adjustment for the change of variables.

The model block defines the log probability function on the
constrained parameter space.  The example in
Figure~\ref{hier-binom.fig} has a simple model containing four
sampling statements.  The hyperprior on the prior mean \code{lambda}
is uniform, and the hyperprior on the prior count \code{kappa} is a
Pareto distribution with lower-bound of support at 0.1 and shape 1.5,
leading to a probability of $\kappa > 0.1$ proportional to
$\kappa^{-5/2}$.  Note that the hierarchical prior on \code{theta} is
vectorized: each element of \code{theta} is drawn independently from
a beta distribution with prior success count \code{alpha} and prior
failure count \code{beta}.  Both \code{alpha} and \code{beta} are
transformed parameters, but because they are only used on the
right-hand side of a sampling statement do not require a Jacobian
adjustment of their own.  The likelihood function is also vectorized,
with the effect that each success count \code{y[i]} is drawn from a
binomial distribution with number of trials \code{n[i]} and chance of
success \code{theta[i]}.  In vectorized sampling statements, single
values may be repeated as many times as necessary.

The (optional) generated quantities allows values that depend on
parameters and data, but do not affect estimation, to be defined
efficiently.  The generated quantities block is called only once per
sample, not once per log probability function evaluation.  It may be
used to calculate predictive inferences as well as to carry out
forward simulation for predictive posterior checks; see
\citep{GelmanEtAl:2013} for examples.

Posterior comparisons can be carried out directly or using rankings.
For instance, the model in Figure~\ref{hier-binom.fig} sets
\code{highest[j]} to 1 if hospital \code{j} has the highest estimated
mortality rate.  Applying a hierarchical model then considering
posterior inference appropriately adjusts for mulitple comparisons;
see \citep{GelmanEtAl:2012, Efron:2010} for discussion.

\proglang{Stan}'s samplers and optimizers all start from either random
or user-supplied values for each parameter.  User supplied initial
values are validated and transformed to the underlying unconstrained
space; if a parameter value does not satisfy its declared constraints,
the program exits and an informative error message is printed.  If
random initialization is specified, the built-in pseudorandom number
generator is called once per unconstrained variable dimension.  The
default initialization is to randomly generate values uniformly on
$[-2,2]$ by default or another interval by specification.  This
supplies fairly diffuse starting points when transformed back to the
constrained scale, and thus help with convergence diagnostics as
discussed in \citep{GelmanEtAl:2013}.  Models with more data or more
elaborate structure require narrower intervals for initialization to
ensure the sampler is able to quickly converge to a stationary
distribution in the high mass region of the posterior.

\subsection{Variable definition and block execution summary}

A table summarizing the point at which variables are read, written,
or defined is provided in Figure~\ref{block-actions.fig}. 
%
\begin{figure}
\begin{center}
\begin{tabular}{l|c|l|l}
{ Block} & { Statements?} & { Action} & { Evaluated}
\\\hline\hline
\code{user initialization} & n/a & transform & chain
\\[3pt]
\code{random initialization} & n/a & randomize & chain 
\\\hline\hline
\code{data} & no & read & chain  
\\
\code{transformed data} & yes & evaluate & chain  
\\ \hline
\code{parameters} & no & inv.\ transform, Jacobian & leapfrog  \\
& & inv.\ transform, write & sample 
\\[3pt]
\code{transformed parameters} & yes & evaluate & leapfrog \\
& & write & sample 
\\\hline
\code{model} & yes & evaluate & leapfrog
\\\hline
\code{generated quantities} & yes & evaluate & sample \\
& & write & sample
\end{tabular}
\end{center}
\caption{\it Each \proglang{Stan} program block admits certain actions that are evaluated
  at specific times during sampling.  For example, the parameter initialization in
  the \code{data} block requires a read operation once per chain.}
\label{block-actions.fig}
\end{figure}
%
This table is defined assuming HMC or NUTS samplers, which require a
log probability and gradient calculation for one or more leapfrog
steps per iteration; see \citep{HoffmanGelman:2011}.
%
\begin{figure}
\begin{center}
\begin{tabular}{l|l}
{ Variable Categorization} & { Declaration Block}
\\ \hline\hline
% constants & \code{transformed data}
% \\ \hline
unmodeled data & \code{data}, \code{transformed data}
\\ 
modeled data & \code{data}, \code{transformed data}
\\ \hline
missing data & \code{parameters}, \code{transformed parameters}
\\
modeled parameters & \code{parameters}, \code{transformed parameters}
\\
unmodeled parameters & \code{data}, \code{transformed data}
\\[2pt] \hline
generated quantities & \code{transformed data}, \code{transformed parameters}, 
\\ 
& \code{generated quantities}
\\ \hline\hline
loop indices & any loop statement
\\ \hline
local variables & any statement block
\\ 
\end{tabular}
\end{center}
\caption{\it Variables of each categorization must be declared in
  specific blocks.  Data may also be expressed using numeric
  literals.}\label{variable-kinds.fig}
\end{figure}
%
\citet[p.~366]{GelmanHill:2007} provide a taxonomy of the kinds of
variables used in Bayesian models.  Figure~\ref{variable-kinds.fig}
contains Gelman and Hill's taxonomy aligned with the corresponding
locations of declarations and definitions in \proglang{Stan}.
Unmodeled data variables includes size constants and regression
predictors.  Modeled data variables include known outcomes or
measurements.  A data variable or constant literal is modeled in a
\proglang{Stan} program if it (or a variable that depends on it)
occurs on the left-hand side of a sampling statement.  Unmodeled
parameters are provided as data, and are either known or fixed to some
value for convenience, a typical use being the parameters of a weakly
informative prior for a parameter.

A modeled parameter is given a distribution by the model (usually
dependent on data and correlated with other parameters).  This can be
done either by placing it (or a variable that depends on it) on the
left-hand side of a sampling statement or by declaring the variable
with a constraint (see Section~\ref{implicit-prior.section}).

Any variable that occurs in a (transformed) data block can also be
provided instead as a constant.  So a user may decide to write \code{y
~ normal(0,1)} or to declare \code{mu} and \code{sigma} as data, and
write \code{y ~ normal(mu,sigma)}.  The latter choice allows the
parameters to be changed without recompiling the model, but requires
them to be specified as part of the input data.

Missing data is a new variable type included here.  In order to
perform inference on missing data, it must be declared as a parameter
and modeled; see \citep{GelmanEtAl:2013} for a discussion of
statistical models of missing data.  Unlike \proglang{BUGS}, data may
not contain a mixture of missing (i.e., \code{NA}) values and numeric
values.  \proglang{Stan} uses the built-in \proglang{C++}
floating-point (and integer) arithmetic, for which there is no
equivalent of \code{NA}.  See \citep{Stan:2013} for strategies for
coding missing data problems in \proglang{Stan}.

Generated quantities include simple transforms of data required for
printing.  For example, the variable of interest might be a standard
deviation, resulting from transforming a precision parameter $\tau$ to
a scale parameter $\sigma = \tau^{-1/2}$.  A non-linearly transformed
variable will have different effective sample sizes and $\hat{R}$
statistics than the variable it is derived from, so it is convenient
to define variables of interest in the generated quantities block to
calculate these statistics automatically.  As seen in the example in
Figure~\ref{hier-binom.fig}, generated quantities may also be used for
event probability estimates.

The generated quantities block may also be used for forward
simulations, generating values to make predictions or to perform
posterior predictive checks; see \citep{GelmanEtAl:2013} for more
information.  The generated quantities block is the only location in
\proglang{Stan} in which random-number generators may be applied
explicitly (they are implicit in parameters).

>>>>

\newenvironment{absolutelynopagebreak}
  {\par\nobreak\vfil\penalty0\vfilneg
   \vtop\bgroup}
  {\par\xdef\tpd{\the\prevdepth}\egroup
   \prevdepth=\tpd}

@@

  \proglang{Stan} can be called from the command line, through
  \proglang{R} using the \pkg{RStan} package, or through
  \proglang{Python} using the \pkg{PyStan} package.  All three
  interfaces support sampling or optimization-based inference and
  analysis, and \pkg{RStan} and \pkg{PyStan} also provide access to
  log probabilities, gradients, Hessians, and data I/O.
}

@@

\section[Introduction]{Introduction}

The goal of the \proglang{Stan} project is to provide a flexible
probabilistic programming language for statistical modeling along with
a suite of inference tools for fitting models that are robust,
scalable, and efficient.

\proglang{Stan} differs from \proglang{BUGS}
\citep{LunnEtAl:2000,LunnEtAl:2009,LunnEtAl:2012} and
\proglang{JAGS}~\citep{Plummer:2003} in two primary ways.  First,
\proglang{Stan} is based on a new imperative probabilistic programming
language that is more flexible and expressive than the declarative
graphical modeling languages underlying \proglang{BUGS} or
\proglang{JAGS}, in ways such as declaring variables with types and
supporting local variables and conditional statements.  Second,
\proglang{Stan}'s Markov chain Monte Carlo (MCMC) techniques are based
on Hamiltonian Monte Carlo (HMC), a more efficient and robust sampler
than Gibbs sampling or Metropolis-Hastings for models with complex
posteriors.%
%
\footnote{\cite{Neal:2011} analyzes the scaling benfit of HMC with
  dimensionality.  \cite{Hoffman-Gelman:2014} provide practical
  comparisions of \proglang{Stan}'s adaptive HMC algorithm with Gibbs,
  Metropolis, and standard HMC samplers.}

\proglang{Stan} has interfaces for the command-line shell
(\pkg{CmdStan}), \proglang{Python} (\pkg{PyStan}), and \proglang{R}
(\pkg{RStan}), and runs on Windows, Mac OS X, and Linux, and is
open-source licensed.

The next section provides an overview of how \proglang{Stan} works by
way of an extended example, after which the details of \proglang{Stan}'s
programming language and inference mechanisms are provided.

\section{Core Functionality}\label{example.section}

@@

Consider estimating the chance of success parameter for a Bernoulli
distribution based on a sequence of observed binary outcomes.  
Figure~\ref{bernoulli-model.fig} provides an implementation of such a
model in \proglang{Stan}.%
%
\footnote{This model is available in the \proglang{Stan} source
  distribution in \nolinkurl{src/models/basic\_estimators/bernoulli.stan}.}
%
\begin{figure}
\begin{Code}
data {
  int<lower=0> N;                  // N >= 0
  int<lower=0,upper=1> y[N];       // y[n] in { 0, 1 }
}
parameters {
  real<lower=0,upper=1> theta;     // theta in [0, 1]
}
model {
  theta ~ beta(1,1);               // prior
  y ~ bernoulli(theta);            // likelihood
}
\end{Code}
\caption{\it Model for estimating a Bernoulli parameter.}\label{bernoulli-model.fig}
\end{figure}
%
The model treats the observed binary data, \code{y[1],...,y[N]}, as
independent and identically distributed, with success probability
\code{theta}.  The vectorized likelihood statement can also be coded
using a loop as in \proglang{BUGS}, although it will run more slowly than the
vectorized form:
%
\begin{absolutelynopagebreak}
\begin{Code}
for (n in 1:N)
  y[n] ~ bernoulli(theta);
\end{Code}
\end{absolutelynopagebreak}
%
A \code{beta(1,1)} (i.e., uniform) prior is placed on \code{theta},
although there is no special behavior for conjugate priors in
\proglang{Stan}.  The prior could be dropped from the model altogether
because parameters start with uniform distributions on their support,
here constrained to be between 0 and 1 in the parameter declaration
for \code{theta}.

@@

Data for running \proglang{Stan} from the command line can be included
in \proglang{R} dump format.  All of the variables declared in the
data block of the \proglang{Stan} program must be defined in the data
file.  For example, 10 observations for the model in
Figure~\ref{bernoulli-model.fig} could be encoded as%
%
\footnote{This data
file is provided with the \proglang{Stan} distrbution in file
\nolinkurl{src/models/basic\_estimators/bernoulli.R.stan}.}
%
\begin{absolutelynopagebreak}
\begin{Code}
N <- 10
y <- c(0,1,0,0,0,0,0,0,0,1)
\end{Code}
\end{absolutelynopagebreak}
%
This defines the contents of two variables, an integer \code{N} and a
10-element integer array \code{y}.  The variable \code{N} is declared
in the data block of the program as being an integer greater than or
equal to zero;  the variable \code{y} is declared as an integer array of size
\code{N} with entries between 0 and 1 inclusive.

@@

The executable can be run with default options by specifying a path to
the data file.  The first command in the following example changes the
current directory to that containing the model, which is where the
data resides and where the executable is built.  From there, the path
to the data is just the file name \code{bernoulli.data.R}.
%
\begin{CodeChunk}
\begin{CodeInput}
> cd $stan/src/models/basic_estimators
> ./bernoulli sample data file=bernoulli.data.R
\end{CodeInput}
\end{CodeChunk}
%
For Windows, the \code{./} before the command should be removed.  This
call specifies that sampling should be performed with the model
instantiated using the data in the specified file.

@@

  Elapsed Time: 0.00932 seconds (Warm-up)
                0.016889 seconds (Sampling)
                0.026209 seconds (Total)
\end{Code}
%
The sampler configuration parameters are echoed, here they are all
default values other than the data file.

The command-line parameters marked \code{Default} may be explicitly
set on the command line.  Each value is preceded by the full path to
it in the hierarchy; for instance, to set the maximum depth for the
no-U-turn sampler, the command would be the following, where backslash
indicates a continued line.
%
\begin{CodeChunk}
\begin{CodeInput}
> ./bernoulli sample  \
  algorithm=hmc engine=nuts max_depth=5  \
  data max_depthfile=bernoulli.data.R
\end{CodeInput}
\end{CodeChunk}
%

@@

The output CSV file, written by default to \code{output.csv}, starts
with a summary of the configuration parameters for the run.
%
\begin{Code}
# stan_version_major = 2
# stan_version_minor = 1
# stan_version_patch = 0
# model = bernoulli_model
# method = sample (Default)
#   sample
#     num_samples = 1000 (Default)
#     num_warmup = 1000 (Default)
#     save_warmup = 0 (Default)
#     thin = 1 (Default)
#     adapt
#       engaged = 1 (Default)
#       gamma = 0.050000000000000003 (Default)
#       delta = 0.80000000000000004 (Default)
#       kappa = 0.75 (Default)
#       t0 = 10 (Default)
#       init_buffer = 75 (Default)
#       term_buffer = 50 (Default)
#       window = 25 (Default)
#     algorithm = hmc (Default)
#       hmc
#         engine = nuts (Default)
#           nuts
#             max_depth = 10 (Default)
#         metric = diag_e (Default)
#         stepsize = 1 (Default)
#         stepsize_jitter = 0 (Default)
# id = 0 (Default)
# data
#   file = bernoulli.data.R
# init = 2 (Default)
# random
#   seed = 847896134
# output
#   file = output.csv (Default)
#   diagnostic_file =  (Default)
#   refresh = 100 (Default)
\end{Code}
%
\proglang{Stan}'s behavior is fully specified by these configuration
parameters, almost all of which have default values.  The
\code{sample} configuration

 By using the same version of
\proglang{Stan} and these configuration parameters, exactly the same
output file can be reproduced.  The pseudorandom numbers generated by
the sampler are fully determined by the seed (here randomly generated
based on the time of the run, with value \code{847896134}) and the
identifier (here \code{0}).  The identifier is used to advance the
underlying pseudorandom number generator a sufficient number of values
that using multiple chains with the same seed and different
identifiers will draw from different subsequences of the pseudorandom
number stream determined by the seed.

The output contiues with a CSV header naming the columns of the
output. For the default NUTS sampler in \proglang{Stan} 2.2.0, these are
%
\begin{Code}
lp__,accept_stat__,stepsize__,treedepth__,n_divergent__,theta
\end{Code}
%
The values headed by \code{lp\_\_} are the log densities (up to an
additive constant), \code{accept\_stat\_\_} are the Metropolis
acceptance proababilities averaged over samples in the slice used by
the no-U-turn sampler, \code{stepsize\_\_} is the leapfrog
integrator's step size for simulating the Hamiltonian,
\code{treedepth\_\_} is the depth of tree explored by the no-U-turn
sampler, and \code{n\_divergent\_\_} is the number of iterations leading
to a numerical instability during integration (e.g., numerical
overflow or a positive-definiteness violation).

 for each iteration.%
%
\footnote{Acceptance is the usual notion for a Metropolis sampler such
  as HMC \citep{MetropolisEtAl:1953}.  For NUTS, the acceptance
  statistic is defined as the average acceptance probabilities of all
  possible samples in the proposed tree; NUTS itself uses a slice
  sampling algorithm for rejection \citep{Neal:2003, HoffmanGelman:2011}.}
%
The column \code{stepsize\_\_} indicates the step size
(i.e., time interval) of the simulated trajectory, while the column 
\code{treedepth\_\_} gives the tree depth for NUTS, defined as the 
log base 2 of the total number of steps in the trajectory.
The rest of the header will be the names of parameters; in this
example, \code{theta} is the only parameter.

@@

\subsection{Roadmap for the Rest of the Paper}

Now that the key functionality of \proglang{Stan} has been
demonstrated, the remaining sections cover specific aspects of
\proglang{Stan}'s architecture.  Section~\ref{data-types.section}
covers variable data type declarations as well as expressions and type
inference, Section~\ref{programming.section} describes the top-level
blocks and execution of a \proglang{Stan} program,
Section~\ref{statements.section} lays out the available statements,
and Section~\ref{functions.section} the built-in math, matrix, and
probability function library.  Section~\ref{inference-engines.section}
lays out MCMC and optimization-based inference.  There are two
appendices, Appedix~\ref{process.section} outlining the development
process and Appendix~\ref{libraries.section} detailing the library
dependencies.

\section{Data types}\label{data-types.section}
